---
title: "FTS state policy"
author: "Pratyoosh Kashyap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Policy Database created using the State FTS Policy Handbook.
-Available at: https://www.farmtoschool.org/resources-main/state-farm-to-school-policy-handbook

Create Policy Environment variables for each state. Categorize them into:
- Policies supporting procurement activities
- Policies supporting education and school garden activities
- Policies which provide other logistical support, promotion programs, infrastructure, etc.

Import libraries and data:

```{r}
library(tidyverse)
library(haven) #to read and write various data formats
library(janitor) #for cleaning data

#import state policy environment
policy <- read_csv("state_fts_policy_2018.csv") %>% clean_names() %>% rename(state_name = state)

#change all NA values to zero:
policy[is.na(policy)] <- 0

#create categories for grouping policies:

#procurement:
policy <- policy %>% mutate(pol_proc = funded_fts + grant + reimbursement_program + local_preference)

#education:
policy <- policy %>% mutate(pol_educ = garden + education)

#other:
policy <- policy %>% mutate(pol_other = fts_coordinator + resolutions + promotion + infrastructure + state_fts + database + pilot + task_force + other_related_policies)

#export dataset:
write_csv(policy, "policy_2018_regdata.csv", na="")
#na="" ensures that NA becomes a blank cell in excel, and not NA. 

```
