---
title: "Econometric Analysis"
author: "Pratyoosh Kashyap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Heckman two-step selection model:
Step 1: Selection Equation = Factors affecting FTS participation (1/0)
Step 2: Treatment Equation = Factors affecting intensity of FTS adoption.

FTS ~ Capitals + Policy + Challenges + SFA + FFVP + RUCC

Import libraries:

```{r}
library(tidyverse)
library(haven) 
library(janitor) 

library(sampleSelection) #to run Heckman 2 step model
library(car) #For VIF
library(ape) #for Moran's I

library(margins) #for marginal effects of regression coefficients

library(stargazer) #to make summary stats table, table of regression output.

```
Import data:

```{r}
#import full data set for regression analysis:
reg_all <- read_csv("reg_data.csv") %>% clean_names()

#Drop all rows which have any NA value:
reg_all <- na.omit(reg_all) #now, total 8162 observations. 

options(scipen=999) # so results are not displayed using scientific notation

```

Heckman model:

```{r}
hman <- heckit(selection = fts ~ factor(ffvp) +
                  pc1b_manufacturing + pc2b_infrastructure + pc1c_artsdiversity + pc2c_creativeindustries + pc1f + pc1h_healtheducation + pc2h_medicalfoodsecurity + pc1n_naturalamenitiesconservation + pc2n_farmland + pc1s_nonprofitsocialindustries + pc2s_publicvoiceparticipation + 
                  pol_proc + pol_educ + pol_other +
                  cvendor + cpurch + ckitch + 
                     log(totalstudents) + poverty_status +
                  factor(rucc_2013),
                  outcome = fts_int ~ pc1b_manufacturing + pc2b_infrastructure + pc1c_artsdiversity + pc2c_creativeindustries + pc1f + pc1h_healtheducation + pc2h_medicalfoodsecurity + pc1n_naturalamenitiesconservation + pc2n_farmland + pc1s_nonprofitsocialindustries + pc2s_publicvoiceparticipation +
                  pol_proc + pol_educ + pol_other +
                  cvendor + cpurch + ckitch + 
                    log(totalstudents) + poverty_status +
                  factor(rucc_2013), 
                 data = reg_all, method = "2step")

summary(hman)

#export

stargazer(hman$probit, hman,
          type = "html",
          style = "aer",
          title = "Regression Results",
          out = "reg.doc",
          intercept.bottom = T,
          intercept.top = F,
          digits = 3,
          single.row = T)

# Marginal Effects for probit:

reg_all <- reg_all %>% mutate(logtotalstudents = log(totalstudents))

probit <- glm(fts ~ factor(ffvp) +
                  pc1b_manufacturing + pc2b_infrastructure + pc1c_artsdiversity + pc2c_creativeindustries + pc1f + pc1h_healtheducation + pc2h_medicalfoodsecurity + pc1n_naturalamenitiesconservation + pc2n_farmland + pc1s_nonprofitsocialindustries + pc2s_publicvoiceparticipation + 
                  pol_proc + pol_educ + pol_other +
                  cvendor + cpurch + ckitch + 
                     logtotalstudents + poverty_status +
                  factor(rucc_2013), 
                  family = binomial(link = "probit"),
                  data = reg_all)

summary(probit)

margins_probit <- summary(margins(probit))

#export marginal effects summary:
write_csv(margins_probit, "margins_probit.csv", na="")

```
Spatial diagnostics - Moran's I:

```{r}
#Define neighbor relationships.

#Using the lat-long coordinates, create a weights matrix:
coordinates <- cbind(reg_all$latitude, reg_all$longitude)

#To calculate an inverse distance weight matrix, generate a matrix of distances (Euclidean):
D <- as.matrix(dist(coordinates))

#Then create a W matrix with each element as the algebraic inverse:
W <- 1/D

#Replace diagonal entries with zero:
diag(W) <- 0

#to see the first 5 rows and columns:
#W[1:5, 1:5]

#change all infinite values to zero.
W[is.infinite(W)] <- 0

#Calculate Moran's I for FTS intensity:
Moran.I(reg_all$fts_int, W) # I = 0.0345

```

Heckman Model with spatial effect originating from FTS intensity:

```{r}
#create spatially weighted fts_int variable:

w_fts_int <- W %*% reg_all$fts_int

#which is a matrix multiplication of nxn * nx1 = nx1.

#head(w_fts_int)

#merge with data set:
reg_sp <- cbind(reg_all, w_fts_int)

#Run Heckman with weighted fts_int in both selection and outcome equations:

hman_sp <- heckit(selection = fts ~ factor(ffvp) +
                  pc1b_manufacturing + pc2b_infrastructure + pc1c_artsdiversity + pc2c_creativeindustries + pc1f + pc1h_healtheducation + pc2h_medicalfoodsecurity + pc1n_naturalamenitiesconservation + pc2n_farmland + pc1s_nonprofitsocialindustries + pc2s_publicvoiceparticipation + 
                  pol_proc + pol_educ + pol_other +
                  cvendor + cpurch + ckitch + 
                    w_fts_int + 
                    log(totalstudents) + poverty_status +
                  factor(rucc_2013),
                  outcome = fts_int ~ pc1b_manufacturing + pc2b_infrastructure + pc1c_artsdiversity + pc2c_creativeindustries + pc1f + pc1h_healtheducation + pc2h_medicalfoodsecurity + pc1n_naturalamenitiesconservation + pc2n_farmland + pc1s_nonprofitsocialindustries + pc2s_publicvoiceparticipation +
                  pol_proc + pol_educ + pol_other +
                  cvendor + cpurch + ckitch + 
                    w_fts_int + 
                    log(totalstudents) + poverty_status +
                  factor(rucc_2013), 
                 data = reg_sp, method = "2step")

summary(hman_sp)

# export regression results:

stargazer(hman_sp$probit, hman_sp,
          type = "html",
          style = "aer",
          title = "Regression Results",
          out = "reg_sp.doc",
          intercept.bottom = T,
          intercept.top = F,
          digits = 3,
          single.row = T)

# Marginal effects:

reg_sp2 <- reg_sp %>% mutate(logtotalstudents = log(totalstudents))

probit_sp <- glm(fts ~ factor(ffvp) +
                  pc1b_manufacturing + pc2b_infrastructure + pc1c_artsdiversity + pc2c_creativeindustries + pc1f + pc1h_healtheducation + pc2h_medicalfoodsecurity + pc1n_naturalamenitiesconservation + pc2n_farmland + pc1s_nonprofitsocialindustries + pc2s_publicvoiceparticipation + 
                  pol_proc + pol_educ + pol_other +
                  cvendor + cpurch + ckitch + 
                    w_fts_int + 
                    logtotalstudents + poverty_status +
                  factor(rucc_2013), 
                  family = binomial(link = "probit"),
                  data = reg_sp2)

summary(probit_sp)

margins_probit_sp <- summary(margins(probit_sp))

#export marginal effects summary:
#write_csv(margins_probit_sp, "margins_probit_sp.csv", na="")

```

Get VIF:

```{r}
ols_m1 <- lm(fts_int ~ pc1b_manufacturing + pc2b_infrastructure + pc1c_artsdiversity + pc2c_creativeindustries + pc1f + pc1h_healtheducation + pc2h_medicalfoodsecurity + pc1n_naturalamenitiesconservation + pc2n_farmland + pc1s_nonprofitsocialindustries + pc2s_publicvoiceparticipation +
                  pol_proc + pol_educ + pol_other +
                  cvendor + cpurch + ckitch +
                  log(totalstudents) + poverty_status +
                  factor(rucc_2013),
                 data = reg_all)

# VIF:

vifm1 <- vif(ols_m1)

#Variance Inflation Factor scores for all of the independent variables were well under 4.0, suggesting little multicollinearity among the variables.

```

